---
title: "UserGuide"
format: html
editor: visual
---

## bulktrends User Guide

Examples of how to use the `bulktrends` package. First, let's load all of the functions contained within the package.

```{r}
#| code-fold: true

library(data.table)
library(ggplot2)
library(devtools)
devtools::load_all("../")
```

## Load UK Trade Info Bulk Data Files

Load a single bulk data file or all of the files in a directory (and subdirectories).

```{r}
if (file.exists("../data/imports.rds")) {
  imports <- readRDS("../data/imports.rds")
} else {
  imports <- read_uktradeinfo( path = "../data/imports/")
  saveRDS(imports, file = "../data/imports.rds")
}

head(imports)

```

## Request Data from UK Trade Data API

Download look-up tables, e.g. for commodity codes and ports, and save, or load if already available.

```{r}
lookup_tables <- c("Commodity","Port")
for ( i in lookup_tables ) {
  
  file <- paste0("../data/",i,".rds")
  
  if( file.exists(file) ){
    assign(i,readRDS(file))
  
    } else {
    assign(i, uktrades_request(endpoint = i)$value)
    saveRDS(get(i), file = file)  
    
    }
}

```

## Quick Visualisation

-   Plot time series of a single commodity

-   Plot the time series of all commodities in "chapter"

```{r}

comcode.plot <- function(code, lookuptable) {
 
  # "code" row from lookup table to meta
  lookuptable <- as.data.table(lookuptable)
  meta <- lookuptable[Cn8Code == code | Hs2Code == code | Hs4Code == code | Hs6Code == code]
  
  #extract related codes once
  allcodes <- na.omit(unique(meta$Cn8Code))
  
  #filter main dataset for all matching codes and merge with meta
  df <- imports[ COMCODE %in% allcodes ]
  df <- merge(df, meta, by.x = "COMCODE", by.y = "Cn8Code", all.x=T)
  
  #add year to df
  df[, year := substr(PERREF, 1, 4)]

  #df[ , .(NET_MASS = sum(as.numeric(NET_MASS))), by = c("year","COMCODE") ]
  
  ## plot year and net mass
  ggplot(df, aes(x = year, y = as.numeric(NET_MASS), fill = factor(COMCODE))) +
  geom_col(show.legend = TRUE) +
  labs(title = paste0("Weight Distribution for ", code, "\n",
                  "Chapter: ", df$Hs2Description[1], " (", df$Hs2Code[1], ")\n",
                  "Heading: ", df$Hs4Description[1], " (", df$Hs4Code[1], ")\n",
                  "Subheading: ", df$Hs6Description[1], " (", df$Hs6Code[1], ")\n",
                  "Commodity Code: ",  df$Cn8LongDescription[1], " (", code, ")\n"),
        x = "Year",
        y = "Weight (Kg)",
        fill = "Commodity Code") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(family= "Arial", face = "bold", size = 10),
        legend.title = element_text(face = "bold"))
}

comcode.plot("01", lookuptable=Commodity)
#comcode.plot("0103", lookuptable=Commodity)
#comcode.plot("01039219", lookuptable=Commodity)


```

-   Hierarchichal classification of a given commodity code

```{r}
comcode.description <- function(code) {
  Commodity <- as.data.table(Commodity)
  meta <- Commodity[
    Commodity$Cn8Code == code |
      Commodity$Hs2Code == code |
      Commodity$Hs4Code == code |
      Commodity$Hs6Code == code,]
  
  desc <- list(
    Hs2 = paste(meta$Hs2Code[1], " — ", meta$Hs2Description[1]),
    Hs4 = paste(meta$Hs4Code[1], " — ", meta$Hs4Description[1]),
    Hs6 = paste(meta$Hs6Code[1], " — ", meta$Hs6Description[1]),
    Cn8 = paste(meta$Cn8Code[1], " — ", meta$Cn8LongDescription[1])
  )
  
  # structured format
  cat("\nHierarchical Description for Code:", code, "\n")
  cat("---------------------------------------------\n")
  cat("Chapter (Hs2):          ", desc$Hs2, "\n")
  cat("Heading (Hs4):            ", desc$Hs4, "\n")
  cat("Subheading (Hs6):         ", desc$Hs6, "\n")
  cat("Commodity code (Cn8):   ", desc$Cn8, "\n")
  cat("---------------------------------------------\n\n")
}

comcode.description("08")
comcode.description("0803")
comcode.description("08039010")
```
