---
title: "UserGuide"
format: html
editor: visual
---

## *bulktrends User Guide*

*Examples of how to use the `bulktrends` package. First, let's load all of the functions contained within the package.*

```{r}
#| code-fold: true
#| warning: false

library(data.table)
library(ggplot2)
library(devtools)
devtools::load_all("../")
```

## *Load UK Trade Info Bulk Data Files*

*Load a single bulk data file or all of the files in a directory (and subdirectories).*

```{r}
if (file.exists("../data/imports.rds")) {
  imports <- readRDS("../data/imports.rds")
} else {
  imports <- read_uktradeinfo( path = "../data/imports/")
  saveRDS(imports, file = "../data/imports.rds")
}

head(imports)

```

## *Request Data from UK Trade Data API*

*Download look-up tables, e.g. for commodity codes and ports, and save, or load if already available.*

```{r}
lookup_tables <- c("Commodity","Port")
for ( i in lookup_tables ) {
  
  file <- paste0("../data/",i,".rds")
  
  if( file.exists(file) ){
    assign(i,readRDS(file))
  
    } else {
    assign(i, uktrades_request(endpoint = i)$value)
    saveRDS(get(i), file = file)  
    
    }
}

```

## *Quick Visualisation*

-   *Plot time series of a single commodity*

-   *Plot the time series of all commodities in "chapter"*

```{r}

#plants
comcode_plot(imports, "06", lookup_table=Commodity, max_unique_comcodes = 10)
comcode_plot(imports, "06029010", lookup_table=Commodity)
#animals
comcode_plot(imports, "01", lookup_table=Commodity, max_unique_comcodes = 10)
comcode_plot(imports, "01012990", lookup_table=Commodity, max_unique_comcodes = 10)

```

-   *Hierarchichal classification of a given commodity code*

```{r}

comcode_description("01012990", Commodity)
comcode_description("08039010", Commodity)
```

-   *Plot monthly volume of imports (2016-2025)*

```{r}

monthly_volume_plot <- function(import_data,
                         code,
                         lookup_table,
                         max_unique_comcodes=10) {
  
  #"code" info to meta
  lookup_table <- as.data.table(lookup_table)
  meta <- lookup_table[Cn8Code == code | Hs2Code == code |
                        Hs4Code == code | Hs6Code == code]
  allcodes <- na.omit(unique(meta$Cn8Code))
  
  #filter main dataset for all matching codes and merge with meta
  import_data <- import_data[ COMCODE %in% allcodes ]
  
  #generate volume
  import_data <- import_data[, .(volume = .N), by=c("COMCODE","month")]

  if ( import_data[,length(unique(COMCODE))] > max_unique_comcodes) {
    topX <- import_data[,.(volume=sum(volume)),by=COMCODE][
      order(volume,decreasing = T),COMCODE][1:max_unique_comcodes]
    import_data[ ! COMCODE %in% topX, COMCODE := "Other"]
    import_data <- import_data[,.(volume=sum(volume)),by=c("COMCODE","month")]
  }
  
  import_data <- merge(import_data , meta,
                     by.x = "COMCODE", by.y = "Cn8Code",
                     all.x=T)
  
  ggplot(import_data, aes(x = month, y = volume, fill = factor(COMCODE))) +
    # geom_col() +
    geom_area() +
    scale_fill_manual(values = colorRampPalette(c("#c6dbef", "#6baed6", "#08306b"))(length(unique(import_data$COMCODE)))) +
    labs(title = "Volume of Imports by Commodity Code",
         subtitle = paste0(
           if (nchar(code) >= 2) {
             import_data[1,paste(Hs2Code, "-", stringr::str_wrap(Hs2Description, width = 85),"\n")]
           } else "",
           if (nchar(code) >= 4) {
             import_data[1,paste(Hs4Code, "-", stringr::str_wrap(Hs4Description, width = 85),"\n")]
           } else "",
           if (nchar(code) >= 6) {
             import_data[1,paste(Hs6Code, "-", stringr::str_wrap(Hs6Description, width = 85),"\n")]
           } else "",
           if (nchar(code) == 8) {
             import_data[1,paste(COMCODE, "-", stringr::str_wrap(Cn8LongDescription, width = 85),"\n")]
           } else ""),
         x = "Year",
         y = "Volume [count/month]",
         fill = paste0("Commodity Code\n(Top ",max_unique_comcodes,")")) +
    guides(fill = guide_legend(keywidth = 0.8, keyheight = 0.6, ncol = 2)) + 
    theme_minimal(base_family = "serif") +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position="top",
          axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(face = "bold", size = 12,
                                    color = "#7D2239", lineheight = 1.1),
          plot.subtitle = element_text(face = "bold", size = 11, color = "#003865"),
          legend.text = element_text(size = 9, family= "serif"),
          legend.title = element_text(face = "bold", size = 11, color = "#003865"))+
    scale_y_continuous(labels = scales::comma)
}
  

  
monthly_volume_plot(imports, "01", Commodity, max_unique_comcodes=10)
monthly_volume_plot(imports, "02012030", Commodity)
monthly_volume_plot(imports, "06", Commodity)



```

-   *Anomaly Detection in time series of a given commodity code*

```{r}

library(forecast)

monthly_volume_outliers <- function(import_data,
                                    code,
                                    lookup_table) {
    
  #"code" info to meta
  lookup_table <- as.data.table(lookup_table)
  meta <- lookup_table[Cn8Code == code | Hs2Code == code |
                       Hs4Code == code | Hs6Code == code]
  allcodes <- na.omit(unique(meta$Cn8Code))
    
  #filter main dataset and generate volume for code
  import_data <- import_data[ COMCODE %in% allcodes ]
  import_data <- import_data[, .(volume = .N), by="month"]
  
  #declare time series
  ts_imports <- ts(import_data$volume, 
                   start = c(2016, 01), 
                   frequency = 12)
  
  #ARX model using forecast::Arima
  trend <- 1:length(ts_imports)
  seas <- fourier(ts_imports, K=1)   #K=1 creates one pair (sin(2πt/12) and cos(2πt/12))
  #X <- model.matrix(~trend + seas)
  X <- cbind(trend, seas)
  
  model <- Arima(ts_imports, 
                 order = c(1,0,0), 
                 xreg = X)
  
  #convert ts data into data.table for ggplot
  res <- data.table(month = import_data$month,
                    fitted = as.numeric(fitted(model)),
                    resid  = as.numeric(residuals(model)))
  
  #find outliers
  threshold <- median(res$resid) + 3*mad(res$resid)
  outlier <- abs(res$resid) > threshold
  
  #plot with threshold
  a <- ggplot(res, aes(x = month, y = resid)) +
       geom_point(aes(colour = outlier), size = 1.2) +
                  scale_colour_manual(values = c("FALSE" = "#003865",
                                                 "TRUE" = "red")) +
       labs(title = paste0("Residual Outliers for " , code),
            x = "Year", 
            y = "Residuals")+
       theme_minimal(base_family = "serif")+
       theme(#panel.grid.major.x = element_blank(),
             #panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12,
                                       colour = "#7D2239", lineheight = 1.1),
             legend.title = element_text(face = "bold", size = 11, colour = "black"))
  
  #plot fitted vs residuals 1
  b <- ggplot(res, aes(x = fitted, y = resid)) +
       geom_point(size=1, colour="#003865") +
       geom_hline(yintercept = 0, linetype = "dashed", colour = "darkgrey")+
       labs(title = paste0("Fitted values vs Residuals for " , code),
            x = "Fitted Values", 
            y = "Residuals")+
       theme_minimal(base_family = "serif")+
       theme(#panel.grid.major.x = element_blank(),
             #panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12,
                                    colour = "#7D2239", lineheight = 1.1),
             legend.title = element_text(face = "bold", size = 11, colour = "black"))
  
  #plot fitted vs residuals 2
  c <- ggplot(res, aes(x = fitted)) +
       geom_line(aes(y = fitted), colour = "#6baed6", linewidth = 0.8) + 
       geom_point(aes(y = resid), colour = "#003865", size = 1) +
       theme_minimal(base_family = "serif") +
       labs(title = paste0("Fitted values vs Residuals for " , code),
            x = "Fitted Values",
            y = "Residuals") +
       theme(#panel.grid.major.x = element_blank(),
             #panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12,
                                       colour = "#7D2239", lineheight = 1.1))
  #plot residuals over time
  d <- ggplot(res, aes(x = month, y = resid)) +
       geom_line(color = "darkgrey") +
       geom_point(size = 1.2, color = "#003865") +
       theme_minimal(base_family = "serif") +
       labs(title = paste0("Residuals Over Time for ", code),
            x = "Year", y = "Residuals")+
       theme(#panel.grid.major.x = element_blank(),
             #panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12,
                                       colour = "#7D2239", lineheight = 1.1))
  
  return(list(model=model, plot=a, plot=b, plot=c,plot=d))
}
    

monthly_volume_outliers(imports, "01", Commodity)
monthly_volume_outliers(imports, "01012100", Commodity)

monthly_volume_outliers(imports, "02", Commodity)
monthly_volume_outliers(imports, "02012030", Commodity)
```
