---
title: "User Guide: `bulktrends`"
format: html
editor: visual
---

```{r install-package}
#| include: false
devtools::install("../", upgrade = FALSE)
```

Examples of how to use the `bulktrends` package (version `r packageVersion("bulktrends")`). First, let's load the package and add the path to where we are storing data.

```{r set-up}
#| warning: false
#| message: false
library(bulktrends)
data_directory <- "../data/"
```

## Load UK Trade Info Bulk Data Files

Load a single bulk data file or all of the files in a directory (and subdirectories). This can take a while for multiple years of data, so it is recommended to save the loaded data as an `.RDS` for quicker loading in the future.

```{r load-trade-data}
#| include: false
if ( !exists("imports") ){
  if (file.exists(paste0(data_directory,"imports.rds"))) {
    imports <- readRDS(paste0(data_directory,"imports.rds"))
  } else {
    imports <- read_uktradeinfo( path = paste0(data_directory,"imports/"))
    saveRDS(imports, file = paste0(data_directory,"imports.rds"))
  }
}
```

```{r}
#| eval: false
imports <- read_uktradeinfo( path = paste0(data_directory,"imports/"))
saveRDS(imports, file = paste0(data_directory,"imports.rds"))
# imports <- readRDS(paste0(data_directory,"imports.rds"))
```

```{r}
head(imports)
```

## Request Data from UK Trade Data API

Download look-up tables, e.g. for commodity codes and ports, and save, or load if already available.

```{r load-lookup-tables}
#| include: false
lookup_tables <- c("Commodity","Port")
for ( i in lookup_tables ) {
  
  file <- paste0(data_directory,i,".rds")
  
  if( file.exists(file) ){
    assign(i,readRDS(file))
  
    } else {
    assign(i, uktrades_request(endpoint = i)$value)
    saveRDS(get(i), file = file)  
    
    }
}
```

```{r}
#| eval: false
Commodity <- uktrades_request(endpoint = "Commodity")$value
```
```{r}
head(Commodity)
```


## Quick Visualisation

The `comcode_plot()` function produces time plots of imports by various measures included in import data, including mass, value (£) and volume (number of imports).

Here are some examples:

```{r plot-examples}
comcode_plot(imports, "01", comcode_lookup=Commodity)

comcode_plot(imports, "0602", variable="STAT_VALUE", comcode_lookup=Commodity)

comcode_plot(imports, "01", variable="volume",comcode_lookup = Commodity)


```

-   Hierarchical classification of a given commodity code

```{r comcode-descriptions}

comcode_description("01012990", Commodity)
comcode_description("08039010", Commodity)
```

-   Detection of outliers in time series of a single commodity

```{r}

# monthly_volume_outliers(imports, "01", Commodity)
# monthly_volume_outliers(imports, "01012100", Commodity)
#monthly_volume_outliers(imports, "01012990", Commodity)
#monthly_volume_outliers(imports, "01061900", Commodity)
#monthly_volume_outliers(imports, "01069000", Commodity)
```
