---
title: "User Guide: `bulktrends`"
format:
  html:
    embed-resources: true
    output-file: index.html
execute:
  freeze: auto
---

```{r install-package}
#| include: false
#| eval: false
try(remove.packages("bulktrends"))
devtools::install("../", upgrade = FALSE)
```

Example usage of the `bulktrends` package (version `r packageVersion("bulktrends")`).

## Set-up

`bulktrends` can be installed from it's [GitHub repository](https://github.com/jbrowell/bulktrends/) using
```{r installation}
#| eval: false
devtools::install_github("jbrowell/bulktrends")
```

Once installed, load the package and set the path to where we are storing data.
```{r set-up}
#| eval: false
library(bulktrends)
data_directory <- "path/to/my_data/"
```
```{r set-up-hidden}
#| warning: false
#| message: false
#| echo: false
library(bulktrends)
data_directory <- "../data/Rbuildignore/"
```

Parallel computation is supported in `bulktrends` using on the [`future` framework](https://future.futureverse.org/). Its use is optional but recommended. To use parallel computation, simply set a "plan". NB: This is currently unreliable, see [Issue #45](https://github.com/jbrowell/bulktrends/issues/45).
```{r future_setup}
#| warning: false
future::plan("sequential")
```
To view progress bars, ask `progressr` to handle them.
```{r progressr_setup}
#| eval: false
progressr::handlers(global = TRUE)
```

## Monthly UK Trade Info Data

### Load UK Trade Info Bulk Data Files

This package has been developed to work with HMRC Monthly Import data. For more details, see the package [README](https://github.com/jbrowell/bulktrends/). We can use it to download monthly trade data from the UK Trade Info and load either a single file or multiple files or all of the files in a directory (and subdirectories). This can take a while for multiple years of data, so it is recommended to save and load this data as a `.gz` file.

```{r load-trade-data-hidden}
#| include: false
if ( !exists("imports") ){
  if (file.exists(file.path(data_directory,"imports.gz"))) {
    imports <- fread(file.path(data_directory,"imports.gz"))
  } else {
    stop("Run script 'data-raw/imports.R' to generate dataset")
  }
}
```

```{r read-trade-data-not-run}
#| eval: false

download_uktradeinfo_bulk(
  dest_dir = file.path(data_directory,"imports/"),
  type = "imports",
  overwrite = FALSE
)
imports <- read_uktradeinfo(path = file.path(data_directory,"imports/"))
fwrite(imports, file = file.path(data_directory,"imports.gz"))
# imports <- fread(file.path(data_directory,"imports.gz"))
```

Alternatively, to update an existing dataset, simply download the latest data using `download_uktradeinfo_bulk()` and then update your dataset using `update_uktradeinfo()`. This function compares the existing dataset with the latest data downloaded and updates it by adding new rows and removing any rows that are no longer present in the latest data. Finally, save the updated dataset as a `.gz` file.

```{r}
#| eval: false
update_uktradeinfo(
  existing = imports,
  path = file.path(data_directory,"imports")
)
fwrite(imports, file = file.path(data_directory,"imports.gz"))
```


```{r head_imports}
#| paged-print: true
head(imports)
```

#### Things to be aware of

-   Country of origin reporting changes from 2022 onward, individual EU countries are specified where previously they were not. There is therefore a large increase in volume (number or rows) of import data (see [methodology and quality report](https://www.gov.uk/government/statistics/overseas-trade-statistics-methodologies/overseas-trade-in-goods-statistics-methodology-and-quality-report--3#data-sources) for more information on the data).

-   Low value Trade EU: Imports via port code QVV only report chapter of `COMCODE` and do not include `NET_MASS`.

### Request Data from UK Trade Data API

Download look-up tables, e.g. for commodity codes and ports, and save, or load if already available.

```{r load-lookup-tables}
#| include: false
lookup_tables <- c("Commodity","Port")
for ( i in lookup_tables ) {
  
  file <- file.path(data_directory,paste0(i,".rds"))
  
  if( file.exists(file) ){
    assign(i,readRDS(file))
    
  } else {
    assign(i, uktrades_request(endpoint = i)$value)
    saveRDS(get(i), file = file)  
    
  }
}
```

```{r}
#| eval: false
Commodity <- uktrades_request(endpoint = "Commodity")$value
```

```{r}
#| paged-print: true
head(Commodity)
```

### Quick Visualisation

The `comcode_plot()` function produces time plots of imports by various measures included in import data, including mass, value (£) and volume (number of imports).

Here are some examples:

```{r plot-examples}
comcode_plot(imports, "01", comcode_lookup=Commodity)

comcode_plot(imports, "0602", variable="STAT_VALUE", comcode_lookup=Commodity)

comcode_plot(imports, "01", variable="volume",comcode_lookup = Commodity)


```

-   Hierarchical classification of a given commodity code

```{r comcode-descriptions}

comcode_description("01012990", Commodity)
comcode_description("08039010", Commodity)
comcode_description("16041428", Commodity)

```

### Filter by commodity groups

`bulktrends` includes useful groupings of commodity codes, such as plants and animals relevant for Sanitary and Phytosanitary Controls (SPS). An object called `comcode_groups` is included in `bulktrends` and loaded along with the package. Warning: `comcode_groups` does not include any suppressed codes, e.g. "01------", take care when using.



```{r}
head(imports[COMCODE %in% comcode_groups$animal_SPS])
```

### Time series extraction

Time series can be extracted for any single or multiple sets of commodity codes. The function can optionally group the data by any available variable, for example `PORT_CODE` for port location and `COO_ALPHA` for country of origin, or a combination of grouping columns. 

When `return_list = TRUE`, the time series are returned separately for each code or combination of code and group. Alternatively, a combined output consisting of the grouping columns, date column, and quantity is returned.

```{r}
#| warning: false

extract_ts(imports, 
           response_col = "NET_MASS",
           code = c("01", "02"), 
           return_list = TRUE)

extract_ts(
  imports,
  response_col = "NET_MASS",
  code = c("01", "02", "03"),
  group_by = c("PORT_CODE"),
  return_list = FALSE
)

extract_ts(
  imports,
  response_col = "volume",
  code = "01",
  group_by = c("PORT_CODE", "COO_ALPHA"),
  return_list = FALSE
)

```


### Anomaly detection: individual

The anomaly detection workflow considers two types of anomalies; structural breaks and temporary outliers.

For a given commodity code and quantity, (i) `extract_ts` extracts a list of time series by summing all commodities with the specified codes by time period, (ii) `select_best_model` automatically finds an optimal regression model from a pool of candidate models of base models and models with structural breaks, and (iii) `detect_outliers` specifically detects temporary outliers ("AO" additive outliers and "TC" temporary changes) using an outlier detection algorithm.

In this workflow, structural break detection is performed within `select_best_model()` by using `break_detection = TRUE`. If any break is identified, the detected break information is stored in the `break_entry` output. 

When `return_tso = TRUE` in `detect_outliers`, the full output object returned by tsoutliers::tso() is retained, allowing users to inspect the detected outliers, model diagnostics, fitted regressors, and temporary outlier effects in detail.


```{r tso_example}
#| warning: false

ts_data <- extract_ts(imports, "01", response_col="NET_MASS")

selected_model <- select_best_model(ts_data, response_col = "NET_MASS", break_detection = TRUE)

detected_anomaly <- detect_outliers(data = selected_model$data,
                    response_col= "NET_MASS",
                    scale_ts = TRUE,
                    return_tso = TRUE,
                    xreg = model.matrix(selected_model$formula, data = selected_model$data),
                    tso_params = list(cval= 4))

print(selected_model$break_entry)
print(detected_anomaly)
plot.tsoutliers(detected_anomaly$tso)


```

### Anomaly detection: large-scale

Detecting anomalies can be performed for a list of commodity codes using the `detect_anomalies()` function. The data to be supplied can either be lists of single time series or a `tsibble` data frame, a time-series table format that supports later analysis. 

The function returns (i) a table of all detected anomalies (breaks and outliers) across all codes, (ii) a `data.table` for each code containing the original time series, the time index, and the associated anomaly covariates and, (iii) the final model formula for each time series.

```{r detect_anomalies_datatable_example}
#| warning: false
#| paged-print: true

hs2_codes <- unique(substr(imports$COMCODE, 1,2))

ts_list <- extract_ts(imports, code = hs2_codes[1:5], return_list = T)

hs2_anomalies <- detect_anomalies(data = ts_list,
                                  response_col = "NET_MASS",
                                  scale_ts = TRUE,
                                  verbose = FALSE,
                                  tso_params = list(cval = 4))

print(hs2_anomalies$outliers)
```


#### Tsibble conversion and anomaly detection

To perform large-scale anomaly detection using data in tsibble format, the function `uktrade_tsibble()` can be used to convert a `data.table` into a `tsibble` data frame. This allows for automatic classification and relevant aggregation according to hierarchies and groups specified. Once in a tsibble format, the data can be used in detect_anomalies() as follows:


```{r detect_anomalies_tsibble_example}
#| warning: false

tsibble_data <- uktrade_tsibble(
  response_col = "NET_MASS",
  data = imports[DATE_START >= "2024-01-01"],
  comcode_level = c("HS2", "HS4"),
  group_by = "category"
)

hs2_anomalies <- detect_anomalies(data = tsibble_data[tsibble_data[["HS2"]] == "01",],
                                  response_col = "NET_MASS",
                                  scale_ts = TRUE,
                                  verbose = FALSE,
                                  tso_params = list(cval = 4, types = c("AO", "TC")))

print(hs2_anomalies$outliers)
```


## Daily IPAFFS Data

### Load IPAFFS Data Files

The Food Standards Agency (FSA) makes some Import of Products, Animals, Food and Feed System (IPAFFS) data public which we can also analyse with `bulktrends`. The data files contain daily data of high-risk products of animal origin and high-risk food of non-animal origin imported into the United Kingdom via approved designated ports (see the [FSA](https://www.food.gov.uk/our-work/imports-intelligence-hub) website for more information). We can load a single csv file or all csv files in a directory (and subdirectories) using `read_ipaffs()`.

```{r load-ipaffs-data}
#| warning: false
imports_ipaffs <- read_ipaffs(path = paste0(data_directory,"imports_fsa_ipaffs"))

```

For consistency, some column names have been standardised. For example, (i) date columns include `DATE_START` and `DATE_END`, (ii) all variations of net weight columns across `.csv` files have been renamed as `NET_MASS` and (iii) all variations of commodity code columns are renamed as `COMCODE`.

```{r}
head(imports_ipaffs[,c(1:4,9,10,55,56,43,17)])
```

### Data Visualisation

The `comcode_plot()` function can be used to also visualise daily time series. For the IPAFFS data, we can currently view `NET_MASS` and `volume` as follows:

```{r}
#| warning: false

comcode_plot(imports_ipaffs[DATE_START>="2024-01-01"],
             code =  "02",
             variable = "NET_MASS",
             comcode_lookup=Commodity)

comcode_plot(imports_ipaffs[DATE_START>="2024-01-01"],
             code =  "02",
             variable = "volume",
             comcode_lookup=Commodity)


ts_data <- extract_ts(imports_ipaffs,
                      code = "15",
                      response_col = "NET_MASS",
                      date_col = "DATE_START",
                      fill_missing = 0,
                      freq = "day")
plot(ts_data)

```

### Anomaly detection: individual

The anomaly detection procedure described using the monthly HMRC data can also be applied to the daily IPAFFS data. This involves using the same functions, such as, `extract_ts` to obtain the time series of a given commodity code, `select_best_model` to identify structural breaks and select an appropriate model from the same set of candidate models, and using `detect_outliers` to identify the various types of temporary outliers.

```{r}
#| eval: true
#| warning: false

ts_data <- extract_ts(imports_ipaffs[DATE_START>="2023-01-01"],
                      code = "07",
                      response_col = "NET_MASS",
                      date_col = "DATE_START",
                      fill_missing=0)

selected_model <- select_best_model(ts_data, response_col = "NET_MASS", break_detection = TRUE)

detected_anomaly <- detect_outliers(data = selected_model$data,
                    response_col= "NET_MASS",
                    scale_ts = FALSE,
                    return_tso = TRUE,
                    xreg = model.matrix(selected_model$formula, data = selected_model$data),
                    tso_params = list(cval= 4))

print(selected_model$break_entry)
print(detected_anomaly)
plot.tsoutliers(detected_anomaly$tso)


```

### Anomaly detection: large-scale

Here is also an example of using the `detect_anomalies()` function to a selected group of commodity codes from the IPAFFS dataset for large-scale anomaly detection. Based on IPAFFS data availability, the function can skip time series that are too short or incomplete. A verbose option is also provided, when set to `TRUE`, progress messages are displayed to help track execution steps. 

Similarly, the function works with both lists of time series and `tsibble` data format, and also returns (i) a table of all detected anomalies (breaks and outliers) across all codes, (ii) a `data.table` for each code containing the original time series, the time index, and the associated anomaly covariates and, (iii) the final model formula for each time series.

```{r}
#| eval: false
#| warning: false

hs2_codes <- sort(unique(substr(imports_ipaffs$COMCODE, 1,2)))

netmass_list <- extract_ts(imports_ipaffs[DATE_START>="2023-01-01"],
                           response_col = "NET_MASS",
                           code = hs2_codes[1:5], 
                           return_list = T)

hs2_anomalies_netmass <- detect_anomalies(data = netmass_list,
                                  response_col = "NET_MASS",
                                  scale_ts = TRUE,
                                  verbose = FALSE,
                                  tso_params = list(cval= 4))


volume_list <- extract_ts(imports_ipaffs[DATE_START>="2023-01-01"],
                          response_col = "volume",
                          code = hs2_codes[1:5], 
                          return_list = T)

hs2_anomalies_volume <- detect_anomalies(data = volume_list,
                                  response_col = "volume",
                                  scale_ts = TRUE,
                                  verbose = FALSE,
                                  tso_params = list(cval= 4))

head(hs2_anomalies_netmass$outliers)
head(hs2_anomalies_volume$outliers)

```

Note: The above anomaly detection procedures for daily data work but can take a while to run so their output has been excluded.
