---
title: "User Guide: `bulktrends`"
format:
  html:
    embed-resources: true
    output-file: index.html
execute:
  freeze: auto
---

```{r install-package}
#| include: false
#| eval: false
try(remove.packages("bulktrends"))
devtools::install("../", upgrade = FALSE)
```

Example usage of the `bulktrends` package (version `r packageVersion("bulktrends")`).

## Set-up

`bulktrends` can be installed from it's [GitHub repository](https://github.com/jbrowell/bulktrends/) using
```{r installation}
#| eval: false
devtools::install_github("jbrowell/bulktrends")
```

Once installed, load the package and set the path to where we are storing data.
```{r set-up}
#| eval: false
library(bulktrends)
data_directory <- "path/to/my_data/"
```
```{r set-up-hidden}
#| warning: false
#| message: false
#| echo: false
library(bulktrends)
data_directory <- "../data/Rbuildignore/"
```

Parallel computation is supported in `bullktrends` using on the [`future` framework](https://future.futureverse.org/). Its use is optional but recommended. To use parallel computation, simply set a "plan".
```{r future_setup}
future::plan("multicore")
```
To view progress bars, ask `progressr` to handle them.
```{r progressr_setup}
#| eval: false
progressr::handlers(global = TRUE)
```

## Monthly UK Trade Info Data

### Load UK Trade Info Bulk Data Files

This package has been developed to work with HMRC Monthly Import data. For more details, see the package [README](https://github.com/jbrowell/bulktrends/). We can use it to load a single bulk data file or all of the files in a directory (and subdirectories). This can take a while for multiple years of data, so it is recommended to save the loaded data as an `.rds` for quicker loading in the future.

```{r load-trade-data-hidden}
#| include: false
if ( !exists("imports") ){
  if (file.exists(paste0(data_directory,"imports.rds"))) {
    imports <- readRDS(paste0(data_directory,"imports.rds"))
  } else {
    imports <- read_uktradeinfo(path = paste0(data_directory,"imports/"))
    saveRDS(imports, file = paste0(data_directory,"imports.rds"))
  }
}
```

```{r read-trade-data-not-run}
#| eval: false
imports <- read_uktradeinfo(path = paste0(data_directory,"imports/"))
saveRDS(imports, file = paste0(data_directory,"imports.rds"))
# imports <- readRDS(paste0(data_directory,"imports.rds"))
```


```{r head_imports}
#| paged-print: true
head(imports)
```

#### Things to be aware of

-   Country of origin reporting changes from 2022 onward, individual EU countries are specified where previously they were not. There is therefore a large increase in volume (number or rows) of import data (see [methodology and quality report](https://www.gov.uk/government/statistics/overseas-trade-statistics-methodologies/overseas-trade-in-goods-statistics-methodology-and-quality-report--3#data-sources) for more information on the data).

-   Low value Trade EU: Imports via port code QVV only report chapter of `COMCODE` and do not include `NET_MASS`.

### Request Data from UK Trade Data API

Download look-up tables, e.g. for commodity codes and ports, and save, or load if already available.

```{r load-lookup-tables}
#| include: false
lookup_tables <- c("Commodity","Port")
for ( i in lookup_tables ) {
  
  file <- paste0(data_directory,i,".rds")
  
  if( file.exists(file) ){
    assign(i,readRDS(file))
    
  } else {
    assign(i, uktrades_request(endpoint = i)$value)
    saveRDS(get(i), file = file)  
    
  }
}
```

```{r}
#| eval: false
Commodity <- uktrades_request(endpoint = "Commodity")$value
```

```{r}
#| paged-print: true
head(Commodity)
```

### Quick Visualisation

The `comcode_plot()` function produces time plots of imports by various measures included in import data, including mass, value (£) and volume (number of imports).

Here are some examples:

```{r plot-examples}
comcode_plot(imports, "01", comcode_lookup=Commodity)

comcode_plot(imports, "0602", variable="STAT_VALUE", comcode_lookup=Commodity)

comcode_plot(imports, "01", variable="volume",comcode_lookup = Commodity)


```

-   Hierarchical classification of a given commodity code

```{r comcode-descriptions}

comcode_description("01012990", Commodity)
comcode_description("08039010", Commodity)
comcode_description("16041428", Commodity)

```

### Filter by commodity groups

`bulktrends` includes useful groupings of commodity codes, such as plants and animals relevant for Sanitary and Phytosanitary Controls (SPS). An object called `comcode_groups` is included in `bulktrends` and loaded along with the package. Warning: `comcode_groups` does not include any suppressed codes, e.g. "01------", take care when using.



```{r}
head(imports[COMCODE %in% comcode_groups$animal_SPS])
```

### Anomaly detection: individual

The anomaly detection workflow considers two types of anomalies; structural breaks and temporary outliers.

For a given commodity code and quantity, a series of helper functions have been created; (i) `extract_ts` to extract an aggregated time series by summing all commodities with the specified code by time period, (ii) `select_best_model` to automatically find an optimal regression model from a pool of candidate models of base models and models with structural breaks, and (iii) `detect_outliers` to specifically detect temporary outliers ("AO" additive outliers, "TC" temporary changes, "IO" innovative outliers) using an outlier detection algorithm.

In this workflow, structural break detection is performed within `select_best_model()` by using `break_detection = TRUE`. If any break is identified, the detected break information is stored in the `break_entry` output. 

When `return_tso = TRUE` in `detect_outliers`, the full output object returned by tsoutliers::tso() is retained, allowing users to inspect the detected outliers, model diagnostics, fitted regressors, and temporary outlier effects in detail.


```{r tso_example}
#| warning: false

ts_data <- extract_ts(imports, "03", quantity="NET_MASS")

selected_model <- select_best_model(ts_data, response_col = "NET_MASS", break_detection = TRUE)

detect_anomaly <- detect_outliers(data = selected_model$data,
                    quantity= "NET_MASS",
                    scale_ts = FALSE,
                    return_tso = TRUE,
                    xreg = model.matrix(selected_model$formula, data = selected_model$data))

print(selected_model$break_entry)
print(detect_anomaly)
plot.tsoutliers(detect_anomaly$tso)


```

### Anomaly detection: large-scale

This procedure can be performed for a list of commodity codes using the `detect_anomalies()` function. It returns (i) a table of all detected anomalies (breaks and outliers) across all codes (ii) a table for each code containing the time index, the original time series and the associated anomaly covariates.

```{r detect_anomalies_example}
#| warning: false
#| paged-print: true

hs2_codes <- unique(substr(imports$COMCODE, 1,2))


hs2_anomalies <- detect_anomalies(import_data = imports,
                                  codes = hs2_codes[1:5],
                                  quantity = "NET_MASS",
                                  scale= TRUE,
                                  freq = "month",
                                  verbose = FALSE,
                                  tso_params = list(cval= 4,
                                                    types = c("AO", "TC", "IO")))

print(hs2_anomalies$outliers)
```

## Daily IPAFFS Data

### Load IPAFFS Data Files

The Food Standards Agency (FSA) makes some Import of Products, Animals, Food and Feed System (IPAFFS) data public which we can also analyse with `bulktrends`. The data files contain daily data of high-risk products of animal origin and high-risk food of non-animal origin imported into the United Kingdom via approved designated ports (see the [FSA](https://www.food.gov.uk/our-work/imports-intelligence-hub) website for more information). We can load a single csv file or all csv files in a directory (and subdirectories) using `read_ipaffs()`.

```{r load-ipaffs-data}
#| warning: false
imports_ipaffs <- read_ipaffs(path = paste0(data_directory,"imports_ipaffs"))

```

For consistency, some column names have been standardised. For example, (i) date columns include `DATE_START` and `DATE_END`, (ii) all variations of net weight columns across `.csv` files have been renamed as `NET_MASS` and (iii) all variations of commodity code columns are renamed as `COMCODE`.

```{r}
head(imports_ipaffs[,c(1:4,9,10,55,56,43,17)])
```

### Data Visualisation

The `comcode_plot()` function can be used to also visualise daily time series. For the IPAFFS data, we can currently view `NET_MASS` and `volume` as follows:

```{r}
#| warning: false

comcode_plot(imports_ipaffs[DATE_START>="2024-01-01"],
             code =  "02",
             variable = "NET_MASS",
             comcode_lookup=Commodity)

comcode_plot(imports_ipaffs[DATE_START>="2024-01-01"],
             code =  "02",
             variable = "volume",
             comcode_lookup=Commodity)


ts_data <- extract_ts(imports_ipaffs,
                      code = "15",
                      quantity = "NET_MASS",
                      date_col = "DATE_START",
                      fill_missing = 0,
                      freq = "day")
plot(ts_data)

```

### Anomaly detection: individual

The anomaly detection procedure described using the monthly HMRC data can also be applied to the daily IPAFFS data. This involves using the same functions, such as, `extract_ts` to obtain the time series of a given commodity code, `select_best_model` to identify structural breaks and select an appropriate model from the same set of candidate models, and using `detect_outliers` to identify the various types of temporary outliers.

```{r}
#| eval: true
#| warning: false

ts_data <- extract_ts(imports_ipaffs[DATE_START>="2023-01-01"],
                      code = "07",
                      quantity = "NET_MASS",
                      date_col = "DATE_START",
                      fill_missing=0)

selected_model <- select_best_model(ts_data, response_col = "NET_MASS", break_detection = TRUE)

detect_anomaly <- detect_outliers(data = selected_model$data,
                    quantity= "NET_MASS",
                    scale_ts = FALSE,
                    return_tso = TRUE,
                    xreg = model.matrix(selected_model$formula, data = selected_model$data))

print(selected_model$break_entry)
print(detect_anomaly)
plot.tsoutliers(detect_anomaly$tso)


```
### Anomaly detection: large-scale

Here is also an example of using the `detect_anomalies()` function to a selected group of commodity codes from the IPAFFS dataset for large-scale anomaly detection. Based on IPAFFS data availability, the function can skip time series that are too short or incomplete. A verbose option is also provided, when set to `TRUE`, progress messages are displayed to help track execution steps. 

Similarly, the function returns (i) a table of all detected anomalies (breaks and temporary outliers) across all codes, and (ii) a table for each code containing the time index, the original time series and the associated anomaly covariates.

```{r}
#| eval: false
#| warning: false

hs2_codes <- sort(unique(substr(imports_ipaffs$COMCODE, 1,2)))

hs2_anomalies_netmass <- detect_anomalies(import_data = imports_ipaffs[DATE_START>="2023-01-01"],
                                  codes = hs2_codes[1:5],
                                  quantity = "NET_MASS",
                                  scale= TRUE,
                                  freq = "day",
                                  verbose = FALSE,
                                  tso_params = list(cval= 4,
                                                    types = c("AO", "TC", "IO")))

hs2_anomalies_volume <- detect_anomalies(import_data = imports_ipaffs,
                                  codes = hs2_codes[1:5],
                                  quantity = "volume",
                                  scale= TRUE,
                                  freq = "day",
                                  verbose = FALSE,
                                  tso_params = list(cval= 4,
                                                    types = c("AO", "TC", "IO")))

head(hs2_anomalies_netmass$outliers)
head(hs2_anomalies_volume$outliers)

```

Note: The above anomaly detection procedures for daily data work but can take a while to run so their output has been excluded.
